Typing a linear \picalc

### 2020-04-13 paper/paper.tex
Slightly shorten the syntax section

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}[theorem]{Definition}
\newtheorem{nitheorem}[theorem]{Theorem}

@@

\subsection{Contribution}

@@

\subsection{Notation}

@@

Abstraction is one of the key reasoning tools in a language: it allows for features to be defined over a range of inputs.
Such is the case of the \picalc{} too, where both scope restriction and input introduce abstractions.
The bodies of these constructs must have a way of referring to their argument.
Using names as variable references is a popular option amongst humans.
However, names are cumbersome to mechanize: inserting a new variable into an environment means proving that the name of such variable is different to all other variable names in the environment.
Moreover, to a machine names are of no significance whatsoever.

Machines prefer things they can algorithmically act upon. Like natural numbers!
What does it mean to use a number as a variable reference?
The idea de Bruijn had \cite{} was to use the index $n$ to refer to the variable introduced $n$ binders ago.
The binders themselves introduce no names anymore.
The expression $\lambda g . (\lambda f . f g) g$ in the \lambdacalc{} would translate as $\lambda (\lambda 0 1) 0$.
That is, terms at different \emph{depths} must use different indices to refer to the same binding.
Humans find this often confusing.
There is however no reason not to keep the original names bestowed by the humans together with the indices.
Machines can then manipulate references mechanically and still use names to present them to humans.

A variable occurring under $n$ abstractions has $n$ things to refer to.
References outside of that range have no associated meaning.
It is useful to rule out these nonsensical terms syntactically.
In Figure \ref{var} we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.

@@

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
As shown in Figure \ref{fig:process}, we increase the \emph{depth} counter every time we create a new channel or receive some input.

@@

Unlike with names, using type-level de Bruijn indices makes our syntax well-scoped by construction.
As a consequence, the semantics of our language can be defined on the totality of the syntax.
User-friendliness can still be recovered through a function that converts processes with names into (possibly) processes with indices.
This function would keep track of what index is associated with what name, and would traverse the process recursively, taking note of new binders and substituting variable references.
If the process is ill-scoped, the function would return nothing.
To print things back to the user names can be substituted with an index together with the name.

@@

\subsection{Structural Congruence}
\label{structural-congruence}

@@

\subsection{Operational Semantics}
\label{operational-semantics}

@@

\subsection{Multiplicities and Capabilities}
\label{multiplicities}

@@

\subsubsection{Example Type Systems}

@@

\subsection{Contexts}
\label{contexts}

@@

\subsection{Typing with Leftovers}
\label{leftover-typing}

@@

\begin{lemma}
  The typing context $\gamma$ at index $i$ is type $t$.
\end{lemma}
\begin{lemma}
  The input annotation in context $\Gamma$ at index $i$ is the result of aggregating (as per the algebra) $y$ and the output annotation in context $\Delta$ at index $i$.
\end{lemma}
\begin{lemma}
  For every index $j$ other than $i$, the input annotation in context $\Gamma$ at index $i$ is the output annotation in $\Delta$ at index $i$.
\end{lemma}
\begin{lemma}
  \todo{Add feedback and feedfront}
\end{lemma}
\begin{corollary}
  Given the witness $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$, $\Gamma \equiv \Delta$ iff $y \equiv \zero$.
\end{corollary}
\proof{All proofs are by structural induction on \textsc{VarRef}.}

@@

\begin{lemma}
\todo{introduce entails-unused}
\end{lemma}

@@

\begin{proof}
\end{proof}

\subsection{Framing}

@@

\subsection{Weakening}

@@

\subsection{Strengthening}

@@

\subsection{Swapping}

@@

\subsection{Subject Congruence}
\label{subject-congruence}

@@

\begin{nitheorem}
  Applying structural congruence rules to a well-typed process should preserve its well-typedness under the same typing and usage contexts.
  If $P \eq{r} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$ then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

@@

\subsection{Substitution}

>>>>

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}[theorem]{Definition}
\newtheorem{nitheorem}[theorem]{Theorem}
\newtheorem{nilemma}[theorem]{Lemma}

@@

\subsection{Contribution}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Notation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In the \picalc{} both scope restriction and input introduce abstractions.
The bodies of these constructs need a way of referring to their argument.
Using names as variable references is a popular option amongst humans.
However, names are cumbersome to mechanize: inserting a new variable into a context means proving that the name of such variable is different from all other variable names in context.
Moreover, names are of no significance to machines.

Machines need constructs they can algorithmically act upon.
The idea de Bruijn had \cite{} was to use a natural number $n$ (aka \emph{index}) to refer to the variable introduced $n$ binders ago.
Binders no longer introduce any names.
\begin{example}
The expression $\lambda g . (\lambda f . f g) g$ in the \lambdacalc{} would translate to $\lambda (\lambda 0 1) 0$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.
Humans often find this confusing.
There is however no reason not to annotate these indices with the original names bestowed by humans.
Machines can then manipulate references mechanically while still using names to present terms to humans.

A variable occurring under $n$ abstractions has $n$ things to refer to.
References outside of that range have no associated meaning.
It is useful to rule out these nonsensical terms syntactically.
In Figure \ref{fig:var} we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.

@@

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
As shown in Figure \ref{fig:syntax}, we increase the \emph{depth} counter every time we create a new channel or receive some input.

@@

Using type-level de Bruijn indices makes our syntax well-scoped by construction, and the semantics of our language can now be defined on the totality of the syntax.
User-friendliness can still be recovered through a function that converts processes with names into (possibly) processes with indices.
This function keeps track of what index is associated with what name, and traverses the process recursively, taking note of new binders and substituting variable references.
If the process is ill-scoped, the function returns nothing.
To present things back to the user, we print indices as their named annotations.

@@

\subsection{Structural Congruence}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{structural-congruence}

@@

\subsection{Operational Semantics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{operational-semantics}

@@

\subsection{Multiplicities and Capabilities}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{multiplicities}

@@

\subsubsection{Example Type Systems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Contexts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{contexts}

@@

\subsection{Typing with Leftovers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{leftover-typing}

@@

\begin{nilemma}
  The typing context $\gamma$ at index $i$ is type $t$.
\end{nilemma}
\begin{nilemma}
  The input annotation in context $\Gamma$ at index $i$ is the result of aggregating (as per the algebra) $y$ and the output annotation in context $\Delta$ at index $i$.
\end{nilemma}
\begin{nilemma}
  For every index $j$ other than $i$, the input annotation in context $\Gamma$ at index $i$ is the output annotation in $\Delta$ at index $i$.
\end{nilemma}
\begin{nilemma}
  \todo{Add feedback and feedfront}
\end{nilemma}
\begin{corollary}
  Given the witness $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$, $\Gamma \equiv \Delta$ iff $y \equiv \zero$.
\end{corollary}
\proof{All proofs are by structural induction on \textsc{VarRef}.}

@@

\begin{nilemma}
\todo{introduce entails-unused}
\end{nilemma}

@@

\begin{nilemma}
  \label{lm:comm-capable}
  \todo{reformulate}
  A well-typed process on which reduction occurs externally, must have this external multiplicity be capable of $\ell_\#$.
\end{nilemma}

\begin{proof}
  For the base case $\constr{comm}$ \todo{}

  Reduction under composition with $\constr{par}$ \todo{}

  Reduction under scope restriction case splits on \textsc{Channel} $c$ on which communication occurs.
  If this channel is internal to the process, or if it is external but on a channel other then the one introduced by the recent scope restriction, subject reduction proceeds inductively.
  However, if reduction happens on the most recently introduced channel, we use the Lemma \ref{lm:comm-capable} to subtract $\ell_\#$ from the channel's multiplicity, then apply the induction hypothesis.

  Finally, in the $\constr{struct}$ case, we apply subject congruence (Theorem \ref{thm:subject-congruence}) and then the induction hypothesis.

\end{proof}

\subsection{Framing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Weakening}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Strengthening}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Swapping}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Subject Congruence}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{subject-congruence}

@@

\begin{nitheorem}
  \label{thm:subject-congruence}
  Applying structural congruence rules to a well-typed process should preserve its well-typedness under the same typing and usage contexts.
  If $P \eq{r} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$ then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

@@

\subsection{Substitution}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
